---
title: "RRR - Replicating Resampling with Rsampling"
author: "Paulo I Prado, Alexandre Oliveira and Andre Chalom"
date: "April 2016"
output:
  rmarkdown::html_vignette:
    fig_width: 5
    fig_height: 5
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{Introduction to Rsampling}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    collapse=TRUE,
    comment = NA,
    prompt = TRUE
    )
set.seed(42)
```

## Overview

This guide is an introduction to the **Rsampling** package, which replicates in R the functions
of the *Resampling Stats* add-in for Excell.
(http://www.resample.com/) [^1].

These functions are used in a workflow that summarizes the logic
behind significance tests:

1. Define a statistic of interest;
2. Define the null hypothesis;
3. Get the statistic of interest distribution under null hypothesis;
4. If the probability of the observed statistic of interest occurring under
null hypothesis is lower than a critical value, reject the null hypothesis.

*Resampling Stats* 's main idea is to facilitate the understanding of this logic,
by making the user execute each step in a spreadsheet, with the aid of
some macros.

**Rsampling** 's package aims at enabling this same training process
in R. Keeping the original workflow is favored over performance.


The sections following installation instructions are examples of the
simpler and most common applications of **Rsampling**. You may refer
to the package help pages to learn about all other functionalities.


## Installation

Rsampling is hosted on CRAN. To install it use

```{r installation CRAN, eval=FALSE}
install.packages("Rsampling")
```
You can also install from the GitHub site, where the project is hosted.
To do that use the devtools package function `install_github`:

```{r installation GitHub, eval=FALSE}
library(devtools)
install_github(repo = 'lageIBUSP/Rsampling')
```

After installation load the package

```{r load library}
library(Rsampling)
```

## Shuffling a column to test difference between groups

The `embauba` data frame contains the data on presence and absence of lianas
on *Cecropia* trees of two morphotypes (white and red).


```{r inspecting object embauba}
head(embauba)
summary(embauba)
```
For more details on the data please refer to
help page (`?embauba`).

### Study hypothesis

The original study's hypothesis is that
ants that live within the hollow trunks of *Cecropia* remove lianas from these trees. 
The prediction is that *Cecropia*  trees of the red morphotype would be less infested by lianas than
the white ones, because they shelter ant colonies more often.
In fact, this difference is observed in the proportions of trees
infested by lianas in the sample:

```{r proportion of infestation by morphotype}
tapply(embauba$with.vines, embauba$morphotype, mean)
```
### Null hypothesis

The null hypothesis is that the proportion of infested trees are equal at
the population from where the samples were drawn.
Under this hypothesis, a liana has the same chance of occurrence in both morphotypes.
We can simulate this null hypothesis
by shuffling the presence of lianas between trees
in the data table.

### Statistic of interest

For each simulation we have to calculate our
**statistic of interest**, which is the
difference of infestation by lianas
between the two morphotypes.
So, we create a function to calculate this statistic:

```{r statistic of interest cecropia}
emb.si <- function(dataframe){
    props <- tapply(dataframe$with.vines, dataframe$morphotype, mean)
    props[[1]] - props[[2]]
}
## Verifying
emb.si(embauba)
```
### Distribution of statistics under the null hypothesis

Then we run the simulation with the function
`Rsampling`:

```{r cecropia resampling, results="hide"}
emb.r <- Rsampling(type = "normal", dataframe = embauba,
                   statistics = emb.si, cols = 2, ntrials = 1000)
```
**What does this command mean?**

* `type = "normal"` indicates a randomization of a whole data set (the most basic type of randomization, afterwards you'll see other types).
* `dataframe = embauba` indicates the data table
* `statistics = emb.si` indicates the function that calculates the statistic of
			interest from the data table.
* `cols = 2` indicates that the randomization must be done over the second column
		of the data table.
* `ntrials = 1000` indicates the number of simulation repetitions.


The distribution of the statistic of interest at the simulation
didn't even include the observed value:

```{r cecropia null distribution, fig.cap="Distribution of the differences between the proportions of red and white *Cecropia* trees with lianas obtained from 1000 simulations of the null hypothesis. The null hypothesis is that the proportion of infested trees are equal at the population from where the samples were drawn. The red line marks the difference observed in the samples and the gray area is the *acceptance region* of the null hypothesis."}
dplot(emb.r, svalue = emb.si(embauba), pside="Greater",
      main = "Frequency distribution of the statistic of interest under H0",
      xlab = "Statistic of interest")
```

### Decision: should we reject the null hypothesis?

As usual in the  biological sciences,
we adopt the criteria of rejecting
the null hypothesis if the probability of
the statistic of interest under the null hypothesis
is under 5% (p < 0.05).

The area not highlighted in gray marks the the top 5%
of the statistic distribution under the null hypothesis.
Thus, if the observed statistic is in the gray area we do not reject
the null hypothesis. This is called the *acceptance region* of H0.
As the observed value (red line) is outside the acceptance region,
H0 can be rejected.
You can also check this with a logical test in
R:

```{r cecropia test} 
sum(emb.r >= emb.si(embauba))/1000 < 0.05
```

**Conclusion:** we reject the null hypothesis (p < 0.05).

## Shuffling within rows to test differences within pairs

The dataframe `azteca` contains the number of  *Azteca* sp ants
recruited by aqueous extracts of old and young leaves of *Cecropia* trees.

```{r inspecting object azteca}
head(azteca)
summary(azteca)
```
Learn more about the data at its help page (`?azteca`).

### Study hypothesis

The study hypothesis is that
recruitment is more intense when a
young leaf is damaged.
This hypothesis predicts that an extract made from young leaves would
recruit more ants than an extract made of old leaves.
Indeed this was the outcome of the experiment:

```{r pairplot azteca, fig.cap = "Number ants that were recruited by aqueous extracts of young and old leaves of *Cecropia* trees. The extracts were dropped in paired leaves of *Cecropia* trees that had colonies of *Azteca* ants. The lines link leaves of same experimental pair."}
splot(azteca$extract.new, azteca$extract.old,
           groups.names=c("Young leaves","Old leaves"),
           ylab="Number of recruited ants",
           xlab="Extract type")
```

### Null hypothesis

The null hypothesis is that the recruitment caused by each extract
is the same. Note that the experiment was paired in order to control for the other sources of variation.
Thus, to simulate the null hypothesis we have to
shuffle the number of recruited ants **within** each pair of
leaves.

### Statistic of interest

For each simulation we calculate our
**statistic of interest**, which is the
mean of the difference of recruited ants within each pair of leaves.
We thus create a  function calculate the statistic of interest from the data table:

```{r statistics of interest azteca}
azt.si <- function(dataframe){
    diferencas <- with(dataframe, extract.new - extract.old)
    mean(diferencas)
}
## Observed value
azt.si(azteca)
```
In the experiment the young leaf extract recruited on average
`r round(azt.si(azteca),1)` ants than the old leaf extract, for each pair.

### Distribution of the statistic under null hypothesis

As the pairs are at rows of our dataframe,
we simulate the null hypothesis shuffling values
within each row:

```{r azteca resampling, results="hide"}
azt.r <- Rsampling(type = "within_rows", dataframe = azteca,
                   statistics = azt.si, cols = 2:3, ntrials = 1000)
```

We changed the argument `type = "within rows"` to indicate that
the values are now shuffled within the lines.
The argument `cols = 2:3` indicates the columns of the dataframe
which contain the counts to be shuffled.

A difference equal to or greater than that observed was very rare
in the distribution of the statistic of interest:

```{r azteca null distribution, fig.cap="Frequency distributions of the difference of the number of ants recruited by the two types of extracts in 1000 simulations of the null hypothesis. The null hypothesis is that both extracts recruit ants equally. The red line marks the difference observed in the samples and the gray area is the *acceptance region* of the null hypothesis."}
dplot(azt.r, svalue = azt.si(azteca), pside="Greater",
      main = "Distribution of the statistic of interest under H0",
      xlab = "Statistic of interest")
```

### Decision: should we reject the null hypothesis?

Again, the distribution of the statistic of interest shows that the observed value of the statistic is outside the acceptance region for the null hypothesis under our significance criterion (5% chance of error).
The same result is found with the following logical test:

```{r azteca test} 
sum(azt.r >= azt.si(azteca))/1000 < 0.05
```

**Conclusion:** we reject the null hypothesis (p<0.05).

#### Coda: one-sided and double-tailed tests.

So far we have tested the hypothesis that a value **equal to or higher** than the observed
can be generated by the null hypothesis. We call this an **one-tailed** or **one-sided** test, as
it would also be if our aim was testing if an equal or smaller value could be generated under null hypothesis.
In one-sided tests, the acceptance region comprises the null distribution except its 5% more extreme values.

But we might investigate differences among groups without specifying its direction. For example,
prior knowledge could suggest the hypothesis that extracts of young and old leaves should recruit
different numbers of ants, but without any expectation concerning which extract would recruit more. This is a case
for a **two-tailed**  or **two-sided** test. In a two-tailed test
the acceptance region is the center of the null distribution, excluding their
2.5% most extreme values at each side:

```{r azteca two-tailed null, fig.cap="Frequency distributions of the difference of the number of ants recruited by the two types of extracts in 1000 simulations of the null hypothesis. The null hypothesis is that both extracts recruit ants equally. The red line marks the difference observed in the samples and the gray area is the *acceptance region* of the null hypothesis for a two-tailed test"}
dplot(azt.r, svalue = azt.si(azteca), pside="Two sided",
      main = "Two-tailed test",
      xlab = "Statistics of interest")
```

## Randomization with replacement

The data frame `peucetia` contains data from an experiment of substrate choice
by spiders of the genus *Peucetia*.
Twenty-seven spiders were kept in Petri dishes
covered with two substrata (plant leaves with and without glandular "hairs" - trichomes - ).
Every plate was inspected six times to check 
if each spider was on the leaves with or without trichomes.

```{r inspecting the peucetia object}
head(peucetia)
```

To learn more about the data  check its help page (`?peucetia`).

### Study hypothesis

The study hypothesis is that
spiders prefer to hunt in plants
with glandular hairs, where catching prey
is easier.
The predicted outcome of the experiment is
that spiders should be recorded most
of the time on leaves with trichomes.
In fact, most of the  spiders were on the leaves with
trichomes in  at least 4 out of six inspections:

```{r barplot peucetia, fig.cap = "Number of inspections each of 27 spiders were recorded on leaves with trichomes in a experiment of preference for two types of substrata (leaves with or without trichomes). The substrate of each spider was checked six times."}
## Number of records in leaves with trichomes
n.insp <- apply(peucetia, 1, sum)
barplot(table(factor(n.insp, levels=0:6)),
        xlab="Number of records in leaves with trichomes",
        ylab="Number of spiders")

```

### Null hypothesis

The null hypothesis is that there is no preference for any substrate.
Half of the area of each plate was covered with each
leaf type. Hence the null expectation
is that the spiders would be
in the area covered by leaves with trichomes in half of the inspections.
This expectation assumes that each inspection
is an independent event. 

### Statistics of interest

For each simulation we have to calculate our
**statistic of interest**. In this case we choose the
average number of inspections where the spiders were recorded on leaves with trichomes.
Thus we build a function to calculate this statistics from data:

```{r statistics of interest peucetia}
peu.si <- function(dataframe){
    mean(apply(dataframe, 1, sum))
}
## Observed value
peu.si(peucetia)
```

The spiders were recorded on average `r round(peu.si(peucetia),2)`
of the six inspections on the area covered with leaves with trichomes.

### Distribution of the statistic under the null hypothesis

To simulate our null hypothesis, we create a
*data frame* with the same structure, wherein each
spider is on leaves with trichomes on half of the inspections.

```{r peucetia H0}
peu.H0 <- matrix( rep(c(TRUE,FALSE), each = 3),
                 nrow = nrow(peucetia), ncol = ncol(peucetia), byrow=TRUE)
## Converts in data.frame
peu.H0 <- data.frame(peu.H0)
## verifying
head(peu.H0)
```
Then we simulate the null hypothesis by sampling each line
with replacement [^2]:

```{r peucetia resampling, results="hide"}
peu.r <- Rsampling(type = "within_rows", dataframe = peu.H0,
                   statistics = peu.si, ntrials = 1000, replace=TRUE)
```

The argument `replace = TRUE` indicates sampling with replacement.
In this case, this is a drawn of an independent position
for each spider at each inspection. Therefore the probability of each spider being at
the area covered by leaves with trichomes is 0.5 per drawing.

An average number of records in leaves with trichomes equal to or greater than that observed 
did not occur at the simulated distribution of our statistic of interest:

```{r peucetia null distribution plot, fig.cap="Frequency distribution of the mean number of records of spiders on leaves with trichomes in 1000 simulations of the null hypothesis that there is no preference for substrata. The red line marks the observed average in the experiment and the gray area is the *acceptance region* of the null hypothesis."}
dplot(peu.r, svalue = peu.si(peucetia), pside="Greater",
      main = "Frequency distribution of the statistics of interest under H0",
      xlab = "Statistics of interest")
```

### Decision: should we reject the null hypothesis?

Again we have a one-tailed test, and the observed value of the statistic of interest fell outside
the region of acceptance of the null hypothesis (5%).
We can check this with the logical test of our significance criterion:

```{r peucetia test} 
sum(peu.r >= peu.si(peucetia))/1000 < 0.05
```

**Conclusion:** we reject the null hypothesis (p < 0.05).

## A more realistic null hypothesis?

In the previous example we simulated the null hypothesis
by drawing a substrate type for each spider for every
inspection. In doing that we assumed that the substrate where a spider is
in an inspection does not affect where the spider will be in the next inspections.
In other words, we assumed that the inspections are
independent events.

But what if the records correlate among inspections?
This can happen if spiders move at a 
smaller frequency than the interval between
inspections. If this is true, subsequent records for a
leaf type may indicate only that spiders tend to stay put,
instead of preference. In this case the null hypothesis
should keep the number of inspections for each leaf type, altering
only the substrata types.

### Null hypothesis

The proportion of inspections that spiders
remain in one of the substrata does not depend on
the substrate type (leaves with or without trichomes).

Therefore the null hypothesis is about the independence between the number
of inspections and type of substrate. We simulate this scenario
by scrambling the number of occasions between substrata,
for each spider. For this we will create a *data frame*
with the observed the number of records of each spider in each substrate:

```{r peucetia n of inspections}
## N of records in leaves with trichomes
tric <- apply(peucetia, 1, sum)
## N of records in leaves w/o trichomes
lisa <- apply(peucetia, 1, function(x) sum(x==0))
## Assembles a data frame with the vectors above
peu.H0b <- data.frame(tric=tric, lisa = lisa)
## Checking 1st lines
head(peu.H0b)
```


### Statistic of interest

A statistic of interest can be applied to
test different null hypotheses. So we keep the same from
our previous example: the average number of inspections where
spiders were recorded on leaves with trichomes.

But since the *data frame* to be randomized
has changed, we create a new function in R to
calculate the statistic of interest

```{r peucetia statistics 2}
peu.si2 <- function(dataframe) mean(dataframe$tric)
## Checking, should be the same a previous calculation
peu.si2(peu.H0b)
```


### Distribution of the statistic of interest under the null hypothesis

In this new case, the null hypothesis is simulated by shuffling the rows
of the *data frame*:

```{r peucetia resampling 2, results="hide"}
peu.r2 <- Rsampling(type = "within_rows", dataframe = peu.H0b,
                   statistics = peu.si2, ntrials = 1000)
```

The null distribution changed significantly when compared to the previous section.
But an average equal to or greater than the observed remained very rare:

```{r peucetia null distribution 2, fig.cap="Frequency distribution of the mean number of records of spiders on leaves with trichomes in 1000 simulations of the null hypothesis that there is no preference for substrata, but taking into account the tendency of spiders to stay put between inspections. The red line marks the observed average in the experiment and the gray area is the *acceptance region* of the null hypothesis."}
dplot(peu.r, svalue = peu.si(peucetia), pside="Greater")
dplot(peu.r2, svalue = peu.si2(peu.H0b), pside="Greater",
      main = "Frequency distribution of the statistic of interest under H0",
      xlab = "Statistic of interest")
```

### Decision: do we reject the null hypothesis?

The observed value of the statistic of interest is not within our acceptance region.
Applying our significance criterion:

```{r peucetia test 2} 
sum(peu.r2 >= peu.si(peucetia))/1000 < 0.05
```

**Conclusion:** we reject the null hypothesis (p < 0.05).


## Structural zeros

Some data sets have observations with zero frequency
that are considered impossible to occur or be observed.
Obvious cases are a cross table of gender and diseases where records
of women with prostate cancer is zero. A less obvious example is in
the *dataframe* `pielou`, which has the number of records of ten
aphid species in twelve species of plants of the genus *Solidago*
 sites across Canada.

```{r pielou inspecting}
pielou
```

To learn more about this data set refer to the help page (`?pielou`).
There are several instances with zero frequency.
We'll simulate a null hypothesis assuming
these frequencies are structural, that is, assuming that
zeroes indicates insect-plant associations that can not occur.
This can be a reasonable assumption for phytophagous insects, that
are in general highly specialized in some host plants.

### Study Hypothesis

Our research hypothesis is that there is or there was resource partitioning of resources among aphid
species. In this case, the observed associations should have resulted in decreased
insect niche overlap.

### Null hypothesis

Our null hypothesis is that the niche overlap does not differ from the expected
if the use plants by plants are independent.

### Statistics of interest

This data was used to illustrate a method to calculate
niche breadth and niche overlap. The expression proposed by the author
for the average niche overlap is the difference between
the Brillouin index of all values and the sum of columns of the data table.
The Brillouin index is a diversity measure in a collection of values $ x_i $:

$$H = \frac{1}{N} \log N! \ - \ \frac{1}{N} \sum \log x_i !$$

Where $N = \sum x_i$. Let's then create a function to calculate this index:

```{r pielou brillouin index}
brillouin <- function(x, base=10) {
    N <- sum(x)
    lfactorial(N)/(log(base)*N)  -  sum(lfactorial(x)/log(base))/N
}
```

Then we create a function to calculate our statistic of interest

```{r pielou statistics}
pielou.si <- function(dataframe)
    brillouin( dataframe ) - brillouin( apply(dataframe,2,sum) )
```
And apply the function to the aphid data:

```{r pielou statistics value}
pielou.si(pielou)
```

### Distribution of the statistic of interest under the null hypothesis

To simulate our null hypothesis, we shuffle the numbers of records
of each species of aphids among the plants. Thus we simulate a situation where we
kept the observed aggregation of records per plant species. But by shuffling records in each
row of the data frame we simulate that these records are independent
Furthermore, we use the `fix.zeroes = TRUE` option to indicate that zero values are not
to be shuffled. In doing this we assume that zeroes indicate associations that can not occur.

```{r pielou randomization, results="hide"}
pielou.r1 <- Rsampling(type = "within_rows", dataframe = pielou,
                   statistics = pielou.si, ntrials = 1000, fix.zeroes = TRUE)
```

The observed value is greater than most values in the null distribution. As our
hypothesis is one-tailed (overlapping observed lower than expected by chance)
the observed value is in the null region of acceptance.

```{r pielou null 2, fig.cap="Frequency distribution of the niche overlap index of aphids in 1000 simulations of the null hypothesis that the associations of aphids and plant species are independent. Zeroes (plants without the record of each species) were kept fixed. The red line shows the observed value of niche overlap."}
dplot(pielou.r1, svalue = pielou.si(pielou), pside="Lesser",
      main = "Frequency distribution of statistics of interest under H0",
      xlab = "Statistics of interest", xlim=c(0.3,0.6))
```

### Decision: do we reject the null hypothesis?

The observed value of our statistic of interest is within the acceptance region.
Applying our significance criterion:

```{r  test 2} 
sum(pielou.r1 <= pielou.si(pielou))/1000 < 0.05
```

**Conclusion:** we do not reject the null hypothesis. (p > 0,05).


[^1]: Statistics.com LCC. 2009. Resampling Stats Add-in for Excel User’s Guide.
http://www.resample.com/content/software/excel/userguide/RSXLHelp.pdf

[^2]: There are simpler and faster ways to do this, but this one replicates
the logic of drawing from an urn in *Rsampling Stats*

